Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-11 ms/cover_letter/plosHIVcover_resubm2.tex
initial values etc.

\newcommand{\revcomment}[1]{\emph{#1}}
\newcommand{\response}[1]{#1}
\newcommand{\rzero}{\ensuremath{\mathcal R}_0}

@@

\textbf{ADD GENERAL POINTS HERE}

@@

The point about initial virulence being measured immediately after
a species jump is interesting, but since it is somewhat peripheral
to our main exploration here, we have opted to include in the 
Discussion rather than trying to squeeze it into the abstract
\textbf{don't forget to do this}
}

@@

\response{
We have changed the wording, and corrected a typo (the first value should have been 0.84 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
}

@@

>>>>

\newcommand{\revcomment}[1]{\emph{#1}}
\newcommand{\response}[1]{#1}
\newcommand{\rzero}{\ensuremath{\mathcal R}_0}
\newcommand{\todo}[1]{{\color \red \bf #1}}

@@

\todo{ADD GENERAL POINTS HERE}

@@

The point about initial virulence being measured immediately after
a species jump is interesting, but since it is somewhat peripheral
to our main exploration here, we have opted to include in the 
Discussion rather than trying to squeeze it into the abstract
\todo{don't forget to do this}
}

@@

\response{
We have changed the wording, and corrected a typo (the second value should have been 0.042 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
}

@@

\response{As part of a sensitivity analysis, to determine the effect of $I(0)$ on the results (since we see that the results are relatively insensitive, we fix a value of $I(0)$ thereafter)
}

### 2017-01-11 ms/ms.tex
initial values etc.

Choosing the initial conditions for the simulations is something of a
challenge. In the most common theoretical modeling contexts
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular quantitative choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
we show results for this value in \figurename~\ref{fig:panel3}.
Since some parameters in our sampling process (see next section)
lead to very low equilibrium prevalence, we switched to 
$I(0)=10^{-4}$ for the remainder of our analysis.
We ran one small set of sensitivity analyses with baseline values of the
epidemiological parameters (Table~1) to test the effects of the
initial conditions. Appendix~S1 provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).

>>>>

Choosing the initial conditions for the simulations is something of a
challenge. In the most common theoretical modeling contexts
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular quantitative choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
we show results for this value in \figurename~\ref{fig:panel3}.
Because we calibrated parameters based on the initial epidemic growth rate
(see next section), we set $I(0)$

lead to very low equilibrium prevalence, we switched to 
$I(0)=10^{-4}$ for the remainder of our analysis.
We ran one small set of sensitivity analyses with baseline values of the
epidemiological parameters (Table~1) to test the effects of the
initial conditions. Appendix~S1 provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As we change $r$ from 0.042 to 0.084 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
